Investigation of the Effect of Pretraining on EEG Classification with Deep Recurrent-Convolutional Networks

### 2017-10-15 paper/paper.tex
intro

\section{Introduction}
\subparagraph{Motivation \& Background} Over the last decades, neural networks have become an extremely popular approach for the data analysis in various fields. Machine learning techniques have proven to play a crucial role in many application areas, because they possess a high classification power, are able to compute high-level abstractions of data, and provide insights into hidden relationships among the data. This is especially important in the field of neuroscience, medical signal and image processing, since the usual real-word uncertainties and incomplete information in the data set are often compounded by a poor understanding of the physical mechanisms by which the data is generated and subjective evaluation of the data by a human observer \cite{roberts}. Electroencephalography (EEG) data analysis relies a lot on statistics, and one of the challenges in modeling cognitive events from EEG data is finding representations that are invariant to inter- and intra-subject differences, therefore, machine learning methods are a crucial tool for EEG-based research. There are many clinical medical applications which include EEG-based brain-computer interface systems, or devices that are intended to change the lives of patients with disabilities, and a lot of research is being done in the area.

Many recent studies \cite{cecotti, shamwell,driver} investigated the application of convolutional neural networks (ConvNets), which are neural networks that can learn local patterns in data by using convolutions as their key component, in the field of EEG analysis. For example, Schirrmeister et al. \cite{schirrmeister} showed how to design and train deep ConvNets with a range of different architectures to achieve the prominent results in EEG decoding tasks, by applying recent advances from the machine learning field and focusing on learning from the raw data of EEG.

Sleep analysis still remains the most unexplored area of neuroscience. Despite the fact that a lot of scientists in the field of sleep studies do not trust innovative automatic approaches, neural networks have been successfully applied to identify different sleep stages \cite{sleep} and their applications for extracting categorical cognitive sleep information are a huge focus of on-going research. Baumgart-Schmitt et al. were able to provide a reasonable agreement of the automatically generated profiles using only one EEG-channel with the profiles scored by sleep-experts on the basis of at least three EEG-channels \cite{somnology}, by applying the system SASCIA (Sleep Analysis System to Challenge Innovative Artificial Networks) to automate the sleep stage scoring.

The state-of-the-art approach to analysing and modeling EEG during sleep recordings was presented by Bashivan et al. in the paper "Learning Representations from EEG with Deep Recurrent-Convolutional Neural Networks" \cite{learning_eeg_repr}. Bashivan et al. used a novel approach, transforming EEG data into a sequence of multi-spectral images which preserve spatial information, which, in its turn, is an important property of EEG methodology. The proposed approach showed significant improvements in accuracy of the performance on the cognitive load classification task.

\section{Data}
\subsection{Data origin \& structure}
\label{ssec:data_structure}
The data set was created at a sleep laboratory \textbf{(ask:the name of the lab?)} in Canada and kindly shared with us for the purpose of this project. The researchers in the laboratory analyse and model brain activity, with the use of Electoencephalography (EEG). EEG is a noninvasive method to record electrical activity of the brain, by placing electrodes along the scalp. We received data from one experiment which was conducted in the laboratory and focused on modeling dream and sleep activity and correlating it with sleep reports. The participants were asked to come to the lab, perform a virtual tennis game, or spatial navigation task, or both and then, directly after that, to go to sleep. After a short period of time, the participants were waken up and asked to verbally describe what they dreamt about. In some session types, the full night sleeps were additionally recorded. For this particular experiment, 22-channel EEG setup was applied, with the sampling rate of 128 samples per second, and the electrodes were customized to the participants according to the International 10-20 system, as can be seen in the figure 1.
\begin{figure}[!ht]
	\centering
	\includegraphics[scale=0.1]{eeg}
	\caption{International 10-20 EEG placement system}
\end{figure}

All of the EEG data we received was already cleaned, so no additional filtering, or artifact rejection was needed to be performed. An example of the EEG data before preprocessing can be seen in figure 2.  For each of the 26 subjects who participated in the experiment \textbf{(look up: how many trials?)}, 3 different groups of data were provided: dream reports, mental rehearsals and verbal reports.  Dream reports contained EEG data during the short naps, mental rehearsals contained EEG data of when the participants were asked to repeat the same task they did in their minds (to think about doing the task without actually doing it). Verbal reports consisted of texts with the description of a dream provided by the participants after waking up. Each instance in the data was annotated with a label of the session type (tennis, spatial navigation, or both).
\begin{figure}[!ht]
	\centering
	\includegraphics[scale=0.2]{eeg_ex}
	\caption{The sample of EEG data from dream reports from one of the participants. EEG traces (horizontal: time in seconds; vertical: channel locations).}
\end{figure}

>>>>

\section{Introduction}
\subparagraph{Motivation} Over the last decades, neural networks have become an extremely popular approach for the data analysis in various fields. An artificial neural network (ANN) is a computational model that is designed to simulate the way the human brain analyzes and processes information. ANNs are based on the structure and functions of biological neural networks and are able to automatically learn and improve from experience without being explicitly programmed. ANNs are considered nonlinear statistical data modeling tools where the complex relationships between inputs and outputs are modeled or patterns are found.

Machine learning (application of ANNs) techniques have proven to play a crucial role in many application areas, because they possess a high classification power, are able to compute high-level abstractions of data, and provide insights into hidden relationships among the data. This is especially important in the field of neuroscience, medical signal and image processing. Often data sets might have incomplete information, which is compounded by a poor understanding of the physical mechanisms by which the data is generated and subjective evaluation of the data by a human observer \cite{roberts}. Electroencephalography (EEG) data analysis relies a lot on statistics, and one of the challenges in modeling cognitive events from EEG data is finding representations that are invariant to inter- and intra-subject differences, therefore, machine learning methods are a crucial tool for EEG-based research. There are many clinical medical applications which include EEG-based brain-computer interface systems, or devices that are intended to change the lives of patients with disabilities, and a lot of research is being done in the area.

\subparagraph{Related work} Many recent studies \cite{cecotti, shamwell,driver} investigated the application of convolutional neural networks (ConvNets), which are neural networks that can learn local patterns in data by using convolutions as their key component, in the field of EEG analysis. In mathematics convolution is an operation on two functions to produce a third function. Each convolutional layer provides a more abstract feature representation, applying a convolution operation to the input, passing the result to the next layer. Convolution here means same filter applied in parallel to all positions in the input. Convolutional layer applies multiple filters to all input positions from layer below.

Schirrmeister et al. \cite{schirrmeister} showed how to design and train deep ConvNets with a range of different architectures to achieve the prominent results in EEG decoding tasks by applying recent advances from the machine learning field and focusing on learning from the raw data of EEG. The architecture desing Schirrmeister et al. used was a  deep ConvNet with four convolution-max-pooling blocks, with a special first block designed to handle EEG input, followed by three standard convolution-max-pooling blocks and a dense softmax classification layer. The input was represented as a 2D-array with the number of time steps as the width and the number of electrodes as the height. This approach significantly reduced the input dimensionality compared with another approach to represent EEG input as a time series of topographically organized images, which was done by Bashivan et al. \cite{learning_eeg_repr}.

Sleep analysis still remains the most unexplored area of neuroscience. One of the bottlenecks in sleep research is the analysis of the EEG data. Many steps, such as identifying the stages of sleep, require visual inspection of the data. Performed by human scientists, scoring and analysis are slow and biased. Neural networks have been successfully applied to identify different sleep stages and their applications for extracting categorical cognitive sleep information are a huge focus of on-going research. Laengvist et al. \cite{sleep} proposed the use of an architecture called deep belief nets (DBNs). A deep belief network is a probabilistic generative model with deep architecture that searches the parameter space by unsupervised greedy layerwise training. Laengvist et al. applied DBNs to the raw EEG input and also validated learned representations by integrating a hidden Markov model to capture a more realistic sleep stage switching, so that the knowledge from the human experts is integrated to the system.

Baumgart-Schmitt et al. were able to provide a reasonable agreement of the automatically generated profiles using only one EEG-channel with the profiles scored by sleep-experts on the basis of at least three EEG-channels \cite{somnology}, by applying the system SASCIA (Sleep Analysis System to Challenge Innovative Artificial Networks) to automate the sleep stage scoring. The core of the SASCIA system consists of neural networks that should separate different sleep stages on the basis of features extracted from only one EEG channel. The features are taken from the input units of the supervised feed-forward network.

The approach to analysing and modeling EEG during sleep recordings, that we based our project on, was presented by Bashivan et al. in the paper "Learning Representations from EEG with Deep Recurrent-Convolutional Neural Networks" \cite{learning_eeg_repr}. Bashivan et al. used a novel approach, transforming EEG data into a sequence of multi-spectral images which preserve spatial information, which, in its turn, is an important property of EEG methodology. The proposed approach showed significant improvements in accuracy of the performance on the cognitive load classification task.

\section{Data}
\subsection{Data origin \& structure}
\label{ssec:data_structure}
The data set was created at the Sleep Research Laboratory (Prof. Stuart Fogel, PhD) at the University of Ottawa in Canada and kindly shared with us for the purpose of this project. The researchers in the laboratory analyse and model brain activity, with the use of Electoencephalography (EEG). EEG is a noninvasive method to record electrical activity of the brain, by placing electrodes along the scalp. We received data from one experiment which was conducted in the laboratory and focused on modeling dream and sleep activity and correlating it with sleep reports. The participants were asked to come to the lab, perform a virtual tennis game, or spatial navigation task, or both and then, directly after that, to go to sleep. After a short period of time, the participants were waken up and asked to verbally describe what they dreamt about. In some session types, the full night sleeps were additionally recorded. For this particular experiment, 22-channel EEG setup was applied, with the sampling rate of 128 samples per second, and the electrodes were customized to the participants according to the International 10-20 system, as can be seen in the figure 1.
\begin{figure}[!ht]
	\centering
	\includegraphics[scale=0.1]{eeg}
	\caption{International 10-20 EEG placement system}
\end{figure}

All of the EEG data we received was already cleaned, so no additional filtering, or artifact rejection was needed to be performed. An example of the EEG data before preprocessing can be seen in figure 2.  For 26 subjects who participated in the experiment there were 766 trials in total, and the experiment consisted of two parts: dream reports and mental rehearsals. Dream reports contained EEG data during the naps, mental rehearsals contained EEG recordings of when the participants were awake and asked to repeat the same task in their minds (to think about doing the task without actually doing it). Verbal reports were recorded for both of above mentioned parts and consisted of texts with the description provided by the participants. Each instance in the data was annotated with a label of the session type (tennis, spatial navigation, or both).
\begin{figure}[!ht]
	\centering
	\includegraphics[scale=0.2]{eeg_ex}
	\caption{The sample of EEG data from dream reports from one of the participants. EEG traces (horizontal: time in seconds; vertical: channel locations).}
\end{figure}
